Bayesian inference of relative fitness on high-throughput pooled competition assays

### 2024-01-20 doc/paper.tex
update corresponding author
TeX file with all journal requirements

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

% Add author affiliations
\usepackage{authblk}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

% Loop through affiliations
\affil[1]{Department of Biology, Stanford University}
\affil[2]{Stanford Cancer Institute, Stanford University School of
Medicine}
\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern
University}
\affil[4]{Department of Molecular Biosciences, Northwestern University}
\affil[5]{Chan Zuckerberg Biohub}

@@

\begin{document}
\maketitle
\begin{abstract}
The tracking of lineage frequencies via DNA barcode sequencing enables
the quantification of microbial fitness. However, experimental noise
coming from biotic and abiotic sources complicates the computation of a
reliable inference. We present a Bayesian pipeline to infer relative
microbial fitness from high-throughput lineage tracking assays. Our
model accounts for multiple sources of noise and propagates
uncertainties throughout all parameters in a systematic way.
Furthermore, using modern variational inference methods based on
automatic differentiation, we are able to scale the inference to a large
number of unique barcodes. We extend this core model to analyze
multi-environment assays, replicate experiments, and barcodes linked to
genotypes. On simulations, our method recovers known parameters within
posterior credible intervals. This work provides a generalizable
Bayesian framework to analyze lineage tracking experiments. The
accompanying open-source software library enables the adoption of
principled statistical methods in experimental evolution.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Remove main text from the table of contents by specifying not to include
% any section or subsection
\addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}

% Define reference segment for main text
\begin{refsegment}
% Generate filter to not include references from main text in the
% supplemental references
\defbibfilter{notother}{not segment=\therefsegment}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The present work is designed to analyze time-series data of relative
abundance of multiple microbial lineages uniquely identified by a DNA
barcode \autocite{kinsler2020,ascensao2023}. In these competition
assays, an ensemble of genotypes is pooled together with an unlabeled
reference strain that, initially, represents the vast majority
(\(\geq 90\%\)) of the cells in the culture (see schematic in
Figure~\ref{fig-01}(A)). Furthermore, a fraction of labeled genotypes
equivalent to the unlabeled reference strain---hereafter defined as
\emph{neutral} lineages---are spiked in at a relatively high abundance
(\(\approx 3-5\%\)). The rest of the culture is left for the ensemble of
genotypes of interest.

@@

\caption{\label{fig-01}\textbf{Typical competitive fitness experiment}.
(A) Schematic of the typical experimental design to determine the
competitive fitness of an ensemble of barcoded genotypes. Genotypes are
pooled together and grown over multiple growth-dilution cycles. At the
end of each cycle, a sample is processed to generate a library for
amplicon sequencing. (B) Typical barcode trajectory dataset. From each
time point, the relative frequency of each barcode is determined from
the total number of reads. Shades of blue represent different relative
fitness. Darker gray lines define the typical trajectory of neutral
lineages.}

@@

To account for all these sources of uncertainty in a principled way, in
the next section, we develop a Bayesian inference pipeline.

@@

The main advantage of our method is this natural interpretability of
these credible regions where an \(X\%\) credible region indeed captures
the region of parameter space where we expect with \(X\%\) probability
the actual value of the parameter lies given our statistical model, our
prior information, and the observed experimental data. A common mistake
in the literature is interpreting frequentist confidence intervals as
Bayesian credible regions when they are not equivalent
\autocite{morey2016}. Frequentist confidence intervals and Bayesian
credible regions are based on fundamentally different philosophical
approaches to statistics. Frequentist confidence intervals represent the
range of values that would contain the true population parameter with a
certain probability if the experiment was repeated many times. The
confidence interval does not represent the probability that the interval
contains the true value. According to a specific model and prior
information, Bayesian credible regions represent the range of values
that contain the parameter with a certain posterior probability. The
credible region directly represents the probability that the region
contains the true value. So, frequentist confidence intervals cannot be
interpreted as Bayesian credible regions because they have fundamentally
different meanings. Treating an \(X\%\) confidence interval like an
\(X\%\) credible region is fallacious since confidence intervals do not
represent probabilistic coverage of the true value like credible
regions. The intervals are generated through entirely different
procedures.

@@

Our inference pipeline can be extended to account for multiple
experimental replicates via Bayesian hierarchical models
\autocite{betancourt2013}. Briefly, when accounting for multiple
repeated measurements of the same phenomena, there are two extreme cases
one can use to perform the data analysis: On the one hand, we can treat
each measurement as entirely independent, losing the power to utilize
multiple measurements when trying to learn a single parameter. This can
negatively impact the inference since, in principle, the value of our
parameter of interest should not depend on the particular experimental
replicate in question. However, this approach does not allow us to
properly ``combine'' the uncertainties in both experiments when
performing the inference. On the other hand, we can pool all data
together and treat our different experiments as a single measurement
with higher coverage. This loses the subtle differences due to biotic
and abiotic batch effects, effectively halving the data that goes into
our inference problem.

@@

Hierarchical models can also capture another experimental design in
which multiple barcodes map to the same or an equivalent genotype. As we
will show, this many-to-one mapping can improve the inference compared
to the extreme cases of inferring the fitness of each barcode
independently or pooling the data of all barcodes mapping to a single
genotype. As schematized in Figure~\ref{fig-06}(A), a small modification
of the base model allows us to map the structure of our original model
to that of a hierarchical model with a fitness hyperparameter vector
\(\underline{\theta}^G\), where \(G\) is the number of genotypes in the
dataset.

@@

It is important to highlight some of the consequences of the general
experimental design and the implicit assumptions within the proposed
statistical model to analyze the resulting data. First, the composition
of the population is such that the initial fraction of the population
occupied by the barcoded genotypes is small---usually \textgreater90\%
of the initial population is the non-labeled reference strain. This
constraint is important as the fitness model used to fit the time series
data assumes that the tracked frequencies are \(\ll 1\). Second, when
computing log frequency ratios, we can run into the issue of dividing by
zero. This is a common problem when dealing with molecular count data
\autocite{lovell2020}. Our model gets around this issue by assuming that
the frequency of any barcode cannot be, but still can get arbitrarily
close to, zero. Therefore, we implicitly assume that no lineage goes
extinct during the experiment. Moreover, the statistical model directly
accounts for the uncertainty associated with having zero barcode counts,
increasing the corresponding uncertainty. Third, the models presented in
this paper require the existence of a labeled sub-population of barcoded
reference strains. These barcodes help determine the fitness baseline,
as every fitness is quantified with respect to this reference genotype.
This experimental design constraint facilitates the inference of the
population mean fitness since most of the culture---the unlabeled
reference genotype---is not tracked. Finally, the presented statistical
model assumes that relative fitness is solely a constant of the
environment and the genotype. Future directions of this work could
extend the fitness model to properly analyze data with time-varying or
frequency-dependent fitness values.

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

We would like to thank Griffin Chure and Michael Betancourt for their
helpful advice and discussion. We would like to thank Karna Gowda,
Spencer Farrell, and Shaili Mathur for critical observations on the
manuscript. This work was supported by the NIH/NIGMS, Genomics of rapid
adaptation in the lab and in the wild R35GM11816506 (MIRA grant), the
NIH, Unravelling mechanisms of tumor suppression in lung cancer
R01CA23434903, the NIH (PQ4), Quantitative and multiplexed analysis of
gene function in cancer in vivo R01CA23125303, the NIH, Genetic
Determinants of Tumor Growth and Drug Sensitivity in EGFR Mutant Lung
Cancer R01CA263715, the NIH, Dissecting the interplay between aging,
genotype, and the microenvironment in lung cancer U01AG077922, the NIH,
Genetic dissection of oncogenic Kras signaling R01CA230025, and the CZ
Biohub investigator program. MRM was supported by the Schmidt Science
Fellowship. MM was supported by The National Science Foundation-Simons
Center for Quantitative Biology at Northwestern University and the
Simons Foundation grant 597491. MM is a Simons Investigator. DP is a CZ
Biohub investigator.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Print main text references
\printbibliography[segment=\therefsegment]
% Close reference segment
\end{refsegment}

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SUPPLEMENTAL MATERIAL

@@

% Define the reference section for the supplemental material
\begin{refsegment}
% Set equation, table, and figure counters to begin with "S"
\beginsupplement
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Add table of contents
\tableofcontents
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% HERE IS WHERE THE SI TEXT GOES
% Print supplemental references changing the title
\printbibliography[title={Supplemental References},
segment=\therefsegment, filter=notother]
\end{refsegment}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

% Add author affiliations
\usepackage{authblk}

% Allow to define personalized colors
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

% Loop through affiliations
\affil[1]{Department of Biology, Stanford University}
\affil[2]{Stanford Cancer Institute, Stanford University School of
Medicine}
\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern
University}
\affil[4]{Department of Engineering Sciences and Applied Mathematics,
Northwestern University}
\affil[5]{Chan Zuckerberg Biohub}

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi

% Remove main text from the table of contents by specifying not to include
% any section or subsection
\addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}

% Define reference segment for main text
\begin{refsegment}
% Generate filter to not include references from main text in the
% supplemental references
\defbibfilter{notother}{not segment=\therefsegment}

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

The tracking of lineage frequencies via DNA barcode sequencing enables
the quantification of microbial fitness. However, experimental noise
coming from biotic and abiotic sources complicates the computation of a
reliable inference. We present a Bayesian pipeline to infer relative
microbial fitness from high-throughput lineage tracking assays. Our
model accounts for multiple sources of noise and propagates
uncertainties throughout all parameters in a systematic way.
Furthermore, using modern variational inference methods based on
automatic differentiation, we are able to scale the inference to a large
number of unique barcodes. We extend this core model to analyze
multi-environment assays, replicate experiments, and barcodes linked to
genotypes. On simulations, our method recovers known parameters within
posterior credible intervals. This work provides a generalizable
Bayesian framework to analyze lineage tracking experiments. The
accompanying open-source software library enables the adoption of
principled statistical methods in experimental evolution.

\hypertarget{author-summary}{%
\section*{Author Summary}\label{author-summary}}
\addcontentsline{toc}{section}{Author Summary}

In this study, we present a novel Bayesian pipeline for analyzing DNA
barcode tracking sequencing data, addressing the challenge of accurately
quantifying competitive microbial fitness in the presence of
experimental noise. Our method uniquely contributes to understanding
microbial evolutionary dynamics by enabling reliable inference of the
relative fitness of diverse microbial strains from high-throughput
lineage tracking assays. Our approach is distinct in its ability to
systematically account for and propagate uncertainties from various
noise sources throughout all inferred parameters. Furthermore, the
error-propagation quality of our Bayesian method allows us to extend the
inference pipeline to common dataset structures, such as jointly
analyzing multiple experimental replicates or accounting for multiple
unique barcodes mapping the equivalent genotypes. This comprehensive
treatment of uncertainties is crucial in experimental settings where
noise can significantly influence the results. Furthermore, we have
optimized our pipeline for scalability, allowing it to handle large
numbers of unique barcodes effectively. This scalability is essential
for analyzing complex datasets typical in microbial fitness studies.

@@

The present work is designed to analyze time-series data of relative
abundance of multiple microbial lineages uniquely identified by a DNA
barcode \autocite{kinsler2020,ascensao2023}. In these competition
assays, an ensemble of genotypes is pooled together with an unlabeled
reference strain that, initially, represents the vast majority
(\(\geq 90\%\)) of the cells in the culture (see schematic in
Figure~\ref{fig-01}(A)). Furthermore, a fraction of labeled genotypes
equivalent to the unlabeled reference strain---hereafter defined as
\emph{neutral} lineages---are spiked in at a relatively high abundance
(\(\approx 3-5\%\)). The rest of the culture is left for the ensemble of
genotypes of interest. This experimental design in which the barcodes of
interest represent a small fraction of the culture serves three main
purposes: First, the model used to infer the relative fitness
differences between genotypes is valid in the regime where the genotype
frequency is significantly smaller than one (see
Section~\ref{sec-fitness_model}). Second, potential problems with
frequency-dependent selection are minimized as long as the frequency of
each genotype remains small. Third, having a single genotype dominating
the culture standardizes the environment experienced by all genotypes.
This is because variations in the chemistry of the environment are
effectively dictated by the dominating genotype, allowing for
reproducible growth-dilution cycles.

@@

\caption{\label{fig-01}\textbf{Typical competitive fitness experiment}.
(A) Schematic of the typical experimental design to determine the
competitive fitness of an ensemble of barcoded genotypes. Genotypes are
pooled together and grown over multiple growth-dilution cycles. At the
end of each cycle, a sample is processed to generate a library for
amplicon sequencing. (B) Typical barcode trajectory dataset. From each
time point, the relative frequency of each barcode is determined from
the total number of reads. Shades of blue represent different relative
fitness. Darker gray lines define the typical trajectory of neutral
lineages. (C) Sources of uncertainty accounted for by our method. The
Bayesian model fit to the data propagates uncertainties from the
categories schematically depicted to all parameters in the inference. }

@@

Figure~\ref{fig-01}(C) shows schematically the sources of uncertainty
accounted for by our model. The first three sources---population
composition, barcode read counts, and barcode frequencies---all
contribute to uncertainty source 1. Uncertainty source 2 and 3 are
depicted by the last two sources, respectively. To account for all these
sources of uncertainty in a principled way, in the next section, we
develop a Bayesian inference pipeline.

@@

The main advantage of our method is the natural interpretability of
these credible regions where an \(X\%\) credible region indeed captures
the region of parameter space where we expect with \(X\%\) probability
the actual value of the parameter lies given our statistical model, our
prior information, and the observed experimental data. Bayesian methods
avoid common misconceptions associated with the construction of
frequentists confidence intervals \autocite{morey2016}.

@@

Our inference pipeline can be extended to account for multiple
experimental replicates via Bayesian hierarchical models
\autocite{betancourt2013}. Briefly, when accounting for multiple
repeated measurements of the same phenomena, there are two extreme cases
one can use to perform the data analysis: On the one hand, we can treat
each measurement as entirely independent, losing the power to utilize
multiple measurements when trying to learn a single parameter. This can
negatively impact the inference since, in principle, the value of our
parameter of interest should not depend on the particular experimental
replicate in question. However, this approach does not allow us to
properly ``combine'' the uncertainties in both experiments when
performing the inference. On the other hand, we can pool all data
together and treat our different experiments as a single measurement
with higher coverage. However, by doing so, we lose all information
about experiment-to-experiment variability due to intrinsic biological
variability and environmental fluctuations---such as small temperature
fluctuations---that we cannot control. In this sense, combining the
datasets defeats the purpose of performing multiple measurements to
account for this variability when extracting the relevant information
from the observations.

@@

Hierarchical models can also capture another experimental design in
which multiple barcodes map to the same or an equivalent genotype. As we
will show, this many-to-one mapping can improve the inference compared
to the extreme cases of inferring the fitness of each barcode
independently or pooling the data of all barcodes mapping to a single
genotype. As schematized in Figure~\ref{fig-06}(A), a small modification
of the base model allows us to map the structure of our original model
to that of a hierarchical model with a fitness hyperparameter vector
\(\underline{\theta}^G\), where \(G\) is the number of genotypes in the
dataset. This analysis pipeline assumes a known barcode-to-genotype
mapping to assign each barcode to the corresponding hyper-fitness
parameter uniquely.

@@

It is important to highlight some of the consequences of the general
experimental design and the implicit assumptions within the proposed
statistical model to analyze the resulting data. First, the composition
of the population is such that the initial fraction of the population
occupied by the barcoded genotypes is small---usually \textgreater90\%
of the initial population is the non-labeled reference strain. This
constraint is important as the fitness model used to fit the time series
data assumes that the tracked frequencies are \(\ll 1\). Second, when
computing log frequency ratios, we can run into the issue of dividing by
zero. This is a common problem when dealing with molecular count data
\autocite{lovell2020}. Our model gets around this issue by assuming that
the frequency of any barcode cannot be, but still can get arbitrarily
close to, zero. Therefore, we implicitly assume that no lineage goes
extinct during the experiment. Moreover, the statistical model directly
accounts for the uncertainty associated with having zero barcode counts,
increasing the corresponding uncertainty. Third, the models presented in
this paper require the existence of a labeled sub-population of barcoded
reference strains. These barcodes help determine the fitness baseline,
as every fitness is quantified with respect to this reference genotype.
This experimental design constraint facilitates the inference of the
population mean fitness since most of the culture---the unlabeled
reference genotype---is not tracked. Fourth, the fitness model in
Equation~\ref{eq-fitness} does not assume any functional form for the
cellular growth dynamics. This description of the changes in the
relative frequency of the genotypes can approximate the dynamics for
multiple growth forms. For example, in Section~\ref{sec-fitseq2}, in the
supplementary materials, we apply our inference pipeline to simulations
like those presented in \autocite{li2023} with non-logistic growth
models. Moreover, Section~\ref{sec-kinsler}, in the supplementary
materials, reanalyzes experimental data from \autocite{kinsler2020},
where previous work showed that the fitness advantage from some of the
mutants results from a shorter lag phase at the beginning of the growth
cycle \autocite{li2018}. Thus, our method's inferred relative fitness is
a coarse-grained quantity that integrates the differences in fitness
across the entire growth cycle as long as frequencies and frequency
changes are small enough to be approximated by
Equation~\ref{eq-fitness}. Finally, the presented statistical model
assumes that relative fitness is solely a constant of the environment
and the genotype. Future directions of this work could extend the
fitness model to properly analyze data with time-varying or
frequency-dependent fitness values.

@@

\hypertarget{materials-and-methods}{%
\section*{Materials and Methods}\label{materials-and-methods}}
\addcontentsline{toc}{section}{Materials and Methods}

All synthetic data generation and custom scripts used in this work were
stored using Git version control. Code for analysis and figure
generation can be found on the GitHub repository
(https://github.com/mrazomej/bayesian\_fitness). The accompanying
software package \texttt{BarBay.jl} can be directly installed from the
\texttt{Julia} package repository, or via cloning the corresponding
GitHub repository (https://github.com/mrazomej/BarBay.jl)

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

We would like to thank Griffin Chure and Michael Betancourt for their
helpful advice and discussion. We would like to thank Karna Gowda,
Spencer Farrell, and Shaili Mathur for critical observations on the
manuscript. We are especially thankful to Grant Kinsler for kindly
providing raw experimental data as well as lengthy discussions about the
state-of-the-art inference method. This work was supported by the
NIH/NIGMS, Genomics of rapid adaptation in the lab and in the wild
R35GM11816506 (MIRA grant), the NIH, Unravelling mechanisms of tumor
suppression in lung cancer R01CA23434903, the NIH (PQ4), Quantitative
and multiplexed analysis of gene function in cancer in vivo
R01CA23125303, the NIH, Genetic Determinants of Tumor Growth and Drug
Sensitivity in EGFR Mutant Lung Cancer R01CA263715, the NIH, Dissecting
the interplay between aging, genotype, and the microenvironment in lung
cancer U01AG077922, the NIH, Genetic dissection of oncogenic Kras
signaling R01CA230025, and the CZ Biohub investigator program. MRM was
supported by the Schmidt Science Fellowship. MM was supported by The
National Science Foundation-Simons Center for Quantitative Biology at
Northwestern University and the Simons Foundation grant 597491. MM is a
Simons Investigator. DP is a CZ Biohub investigator.

% Print main text references
\printbibliography[segment=\therefsegment]
% Close reference segment
\end{refsegment}

\hypertarget{supp_figs}{%
\section*{Supplemental Materials}\label{supp_figs}}
\addcontentsline{toc}{section}{Supplemental Materials}

\textbf{Figure S1. High-dimensional numerical quadrature does not scale with
dimensionality.} Schematic depiction of the problem with naive numerical
quadrature to integrate over an unknown density. While the density is
concentrated on the dark peak, most of the evaluations over the \(x_1-x_2\) grid
do not contribute to the value of the integral.

\textbf{Figure S2. Logistic growth simulation over single growth cycle}. The
dashed line represents the neutral lineages, with the upper curve being the
unlabeled neutral strain. Color curves represent the genotypes of interest
colored by growth rate relative to the neutral lineage.

\textbf{Figure S3. Growth-dilution cycles for logistic growth simulation}. Each
point represents the final number of cells after a growth cycle for each
lineage. Colors are the same as in Figure~\ref{fig-SIX_logistic_growth_01}.

\textbf{Figure S4. Log frequency ratio for logistic growth simulations}. The
relative distance of the color curves from the black dashed line determines the
relative fitness of each linage.

\textbf{Figure S5. Logistic growth-dilution simulations with and without noise}

\textbf{Figure S6. FitSeq2.0 simulation comparison}.(A) Frequency trajectories
that represent the raw data going into the inference. (B) Log frequency ratio
between two adjacent time points used by the inference pipeline. Darker lines
represent the neutral barcodes. These transformed data are much more noisy than
the seemingly smooth frequency trajectories. (C) Examples of the posterior
predictive checks for all neutral lineages (upper left panel) and a subset of
representative mutant lineages. Shaded regions represent the 95\%, 68\%, and 5\%
credible regions for the data. The reported errors above the plot represent the
68\% credible region on the mutant relative fitness marginal distribution. (D)
Comparison between the ground truth fitness value from the logistic-growth
simulation and the inferred fitness value. Gray error bars represent the 68\%
posterior credible region for the relative fitness values. (E) Comparison
between fitness inferences as obtained using \texttt{FitSeq2.0} from
\textcite{li2023} and our Bayesian method. Vertical gray error bars represent
the 68\% posterior credible region for the relative fitness values.

\textbf{Figure S7. Kinsler et al., 2020 sample data}. Barcode frequency
trajectories in four different environments for mutants evolved in minimal
glucose media. Each environment was measured in two biological replicates (R1
and R2 on each panel). For experimental details, see \autocite{kinsler2020}.

\textbf{Figure S8. Posterior predictive checks for Kinsler et al., 2020
independent replicates}. Examples of the posterior predictive checks for all
neutral lineages (upper left panels) and a subset of representative mutant
lineages for all four environments shown in Figure~\ref{fig-SIX_kinsler_data}.
Shaded regions from light to dark represent the 95\%, 68\%, and 5\% credible
regions. Listed fitness values for mutant lineages represent the 68\% credible
region.

\textbf{Figure S9. Replicate-to-replicate correlation for single-dataset model}.
Comparison of inferred relative fitness values for mutant strains between
experimental replicates when fitting each replicate individually for all
environments shown in Figure~\ref{fig-SIX_kinsler_data}. Gray error bars show
the 68\% credible region for this inference.

\textbf{Figure S10. Comparing the replicate-to-replicate deviations with
within-batch noise estimates for single dataset analysis}. Empirical cumulative
distribution functions of the distance between the mean fitness estimate for
each replicate (blue lines) and each replicate's posterior distribution standard
deviation (orange and green line).

\textbf{Figure S11. Posterior predictive checks for Kinsler et al., 2020
hierarchical model}. Examples of the posterior predictive checks for all neutral
lineages (upper left panels) and a subset of representative mutant lineages for
all four environments shown in Figure~\ref{fig-SIX_kinsler_data}. Shaded regions
from light to dark represent the 95\%, 68\%, and 5\% credible regions. Listed
fitness values for mutant lineages represent the 68\% credible region.

\textbf{Figure S12. Replicate-to-replicate correlation for hierarchical model}.
Comparison of inferred fitness values between experimental replicates when
fitting a hierarchical model on the biological replicates for all environments
shown in Figure~\ref{fig-SIX_kinsler_data}. Gray error bars show the 68\%
credible region for this inference.

\textbf{Figure S13. Replicate-to-replicate correlation for both methods}.
Comparison of \textbf{mean} inferred fitness values between experimental
replicates when fitting a single-dataset model (blue points) vs.~a hierarchical
model on the biological replicates (orange points) for all environments shown in
Figure~\ref{fig-SIX_kinsler_data}. Error bars are not included in the plot for
visual clarity.

\textbf{Figure S14. Comparing the replicate-to-replicate deviations with
within-batch noise estimates for hierarchical model analysis}. Empirical
cumulative distribution functions of the distance between the mean fitness
estimate for each replicate (blue lines) and each replicate's posterior
distribution standard deviation (orange and green line).

\textbf{Figure S15. Comparing estimates uncertainties between single-dataset and
hierarchical model}. Comparison of the reported standard deviations for the
mutants' relative fitness when inferring the relative fitness using one dataset
at a time (\(x\)-axis) and the corresponding standard deviation for the mutants'
hyper-fitness parameter when utilizing both replicates simultaneously via a
hierarchical model (\(y\)-axis). Dashed line shows identity line.

\textbf{Figure S16. Scaling of computation time as a function of number of
unique barcodes}. Fitting a single-dataset model for 500 to 100,000 unique
barcodes by taking 7,500 steps.

\clearpage

% SUPPLEMENTAL MATERIAL

@@

% Define the reference section for the supplemental material
\begin{refsegment}
% Set equation, table, and figure counters to begin with "S"
\beginsupplement

% Add table of contents
\tableofcontents

@@

@@

\hypertarget{simulation-details}{%
\subsubsection{Simulation details}\label{simulation-details}}

This section lists the parameters used to generate the synthetic data
used throughout the main text. All the code to reproduce every
simulation and figure in the paper can be found in
\href{https://github.com/mrazomej/bayesian_fitness/tree/master}{the
GitHub repository}. Unless stated otherwise, the simulations were
initialized following these steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The initial number of cells were defined as:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Ancestor lineage: \texttt{n\_init} \(\times\) \texttt{frac\_anc}.
\item
  Neutral lineages:
  \(n_0^{(n)} \sim \text{Poisson} \left(  \texttt{n\_init} \times \frac{\texttt{frac\_neutral}}{\texttt{n\_neutral}} \right)\)
\item
  Mutant lineages:
  \(n_0^{(m)} \sim \text{LogNormal} \left(  \mu = \log(\texttt{n\_init} \times \frac{\texttt{frac\_mut}}{\texttt{n\_mut}}),  \sigma = 2 \right)\)
\end{itemize}

where
